HO unification from object language to meta language

### 2024-03-28 paper/main.tex
w
x

The equational theory of the meta language Elpi that we want to use to
implement a form of proof automation is strikingly similar, since it
it comprises $\eta\beta$ (for the meta language functions), and the
unification procedure \Ue{} solves higher-order problems in
$\llambda$.

@@

We can define $s_1 \Uo{} s_2$ by specializing the code of \hrun{} to
$\mathcal{S} = \{ s_1, s_2 \}$ as follows:
%
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \progress~(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
\quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
\end{array}
$$

@@

\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
\newcommand{\maybebeta}{\ensuremath{\Diamond\beta}\xspace}
\begin{definition}[\maybeeta]
  $\maybeeta = \{ t ~|~ \exists \rho, \rho t ~\mathrm{is~an~eta~expansion} \}$
\end{definition}

@@

\noindent
An example of $t$ in \maybebeta{} is $F a$ for a constant $a$. Note however tha
an oracle could provide an assignment $\rho = \{ F \mapsto \lambda x.x\}$
that makes the resulting term fall outside of \maybebeta.

@@

\begin{proposition}[\wellb{}-preservation]\label{prop:nf}
$\forall \mathcal{T}, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
$$
\wellb(\sigma\mathcal{T}) \land
\stepH{\mathcal{T}}{p}{\sigma}{\linkStore}{\sigma'}
\Rightarrow \wellb(\sigma' \mathcal{T})
$$
\end{proposition}

\noindent
A less formal way to state \ref{prop:nf} is that \hstep{} never
``commits'' an unneeded $\lambda$-abstraction in $\sigma$ (a $\lambda$
that could be erased by an $\eta$-contraction),
nor puts in $\sigma$ a flexible application outside \llambda{} (an application
node that could be erased by a $\beta$-reduction).

Note that proposition \ref{prop:nf} does not hold for \Uo{} since
decompilation can introduce (actually restore) terms in
\maybeeta or \maybebeta that were move out of the way (put in $\linkStore$)
during compilation.

@@

\noindent
The \elpiIn{name} builtin predicate tests if a term is a bound variable.
\footnote{one could always load name x for every x under a pi and get rid of the name builtin}

In both languages unification variables are identified by a natural number
representing a memory address.
The memory and its associated operations are described below:

@@

\noindent
The right hand side of a link, the problematic term, can occur under binders.
To accommodate this situation the compiler wraps \elpiIn{baselink} using
the \elpiIn{inctx} container.

>>>>

The equational theory of the meta language Elpi is strikingly similar,
since it it comprises $\eta\beta$ (for the meta language functions), and the
unification procedure \Ue{} solves higher-order problems in
$\llambda$.

@@

We can define $s_1 \Uo{} s_2$ by specializing the code of \hrun{} to
$\mathcal{S} = \{ s_1, s_2 \}$ as follows:
%
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \progress(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
\quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
\end{array}
$$

@@

\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
\newcommand{\maybebeta}{\ensuremath{\overline{\llambda}}\xspace}
%\newcommand{\maybebeta}{\ensuremath{\Diamond\beta}\xspace}
\begin{definition}[\maybeeta]
  $\maybeeta = \{ t ~|~ \exists \rho, \rho t ~\mathrm{is~an~eta~expansion} \}$
\end{definition}

@@

\noindent
An example of $t$ in \maybebeta{} is $F a$ for a constant $a$. Note however tha
an oracle could provide an assignment $\rho = \{ F \mapsto \lambda x.x\}$
that makes the resulting term fall back in \llambda.

@@

\begin{proposition}[\wellb{}-preservation]\label{prop:nf}
$\forall \mathcal{T}, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
$$
\begin{array}{l}
\wellb(\sigma\mathcal{T}) \land
  \sigma\mathcal{T}_{p_l} \Ue \sigma\mathcal{T}_{p_r} \mapsto {\sigma'}
  \Rightarrow \wellb(\sigma' \mathcal{T})\\
\wellb(\sigma\mathcal{T}) \land
  \progress(\linkStore,\sigma) \mapsto (\_,\sigma')
  \Rightarrow \wellb(\sigma' \mathcal{T})
\end{array}
$$
\end{proposition}

\noindent
A less formal way to state \ref{prop:nf} is that \hstep{} and \progress never
``commit'' an unneeded $\lambda$-abstraction in $\sigma$ (a $\lambda$
that could be erased by an $\eta$-contraction),
nor put in $\sigma$ a flexible application outside \llambda{}
(an application node that could be erased by a $\beta$-reduction).

Note that proposition \ref{prop:nf} does not hold for \Uo{} as a whole
since decompilation can introduce (actually restore) terms in
\maybeeta or \maybebeta that were move out of the way
(put in $\linkStore$) during compilation.

@@

\noindent
In both languages unification variables are identified by a natural number
representing a memory address.
The memory and its associated operations are described below:

@@

\noindent
The right hand side of a link, the problematic term, can occur under binders.
To accommodate this situation the compiler wraps \elpiIn{baselink} using
the \elpiIn{inctx} container (see, \ref{data:inctx}).
